---
title: "Soccer_Analysis"
author: "Adam Saxton"
date: "2026-09-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data loading and preparation

This analysis uses player and team level data from the 2024 English Premier 
League season. The primary objective is to determine if player performance
statistics can distinguish teams that finished in the Top 10 from the teams
that finished in the Bottom 10 at the end of the season.

I used a binary response variable: Success. Success was defined as either a
1 for a player on a Top 10 team, and a 0 for a player on a Bottom 10 team.
Player statistics are used as predictors, teamId was used to ensure that 
players on the same team remained together during validation. This stops
the same team being in both the train and test data sets.
```{r setup-data}
# Libraries
library(readxl)
library(randomForest)
library(dplyr)
library(pROC)
library(caret)
library(ggplot2) 
library(here)



# Load Data
player <- read_excel(here("data", "Soccer_Data.xlsx"), sheet = 1)
team <- read_excel(here("data", "Soccer_Data.xlsx"), sheet = 2)

# Binary Outcome: 1 = Top-10 Finish, 0 = Bottom-10 Finish
player$Success <- factor(ifelse(player$Standing <= 10, 1, 0))
team$Success <- factor(ifelse(team$teamRank <= 10, 1, 0))
```

```{r player-preprocessing}
# Numeric Predictors and keep teamId for validation
columns_to_keep <- c("teamId", "Success", "totalGoals_value", 
                     "goalAssists_value", "yellowCards_value", 
                     "redCards_value", "appearances_value", 
                     "totalShots_value", "shotsOnTarget_value", 
                     "saves_value", "subIns_value", 
                     "foulsCommitted_value", "foulsSuffered_value",
                     "ownGoals_value", "offsides_value", 
                     "goalsConceded_value", "shotsFaced_value")

# Clean Data
player_data_ready <- player %>% select(all_of(columns_to_keep)) %>% 
                     na.omit() %>%
                     mutate(
                       Success = factor(
                        paste0("c", Success), 
                        levels = c("c0", "c1")
                       )
                     )

# One row per team with Top-10 / Bottom-10 label
team_labels <-  player_data_ready %>% distinct(teamId, Success)
```

Incomplete observations have been removed and useful performance variables have
been selected, the data is ready for team-grouped model evaluation. Validation
is performed at the team level because multiple players belong to the same team
and therefore have the same team outcome, so I did not randomly split the 
players.

## Team-Grouped Cross-Validation

To avoid leakage from having players from the same team being in both the
train/test split, five-fold cross-validation is grouped by team. All the players
from each team are grouped together and placed in the same fold, each test
fold contains teams that were not included in model training.

The folds are stratified by Top 10 vs Bottom 10 team status. A Random Forest
classifier is trained within each fold, and performance is measured using player
classification accuracy and ROC AUC. Predicted player probabilities are averaged
within every held-out team to produce out of fold team-level predictions.
```{r grouped-cv-function}
# Stratified team-level folds 
set.seed(123)

team_folds <- team_labels %>% group_by(Success) %>% 
              mutate(fold = sample(rep(1:5, length.out = n())) 
                    ) %>%
              ungroup()

run_grouped_cv <- function(player_data, min_appearances, team_folds) {
  
  fold_results <- data.frame()
  all_team_predictions <- data.frame()
  
  for (k in 1:5){
    
    # Teams assigned to fold become test set
    test_teams <- team_folds %>% filter(fold == k) %>% pull(teamId)
    train_teams <- team_folds %>% filter(fold != k) %>% pull(teamId)
    
    # Appearance cutoff
    filtered_data <- player_data %>% filter(appearances_value >= min_appearances)
    
    
    # Preserve test team IDs before removing teamId from predictors
    test_team_ids <- filtered_data %>%
                     filter(teamId %in% test_teams) %>%
                     pull(teamId)
    
    # Training data: keep all predictor columns, remove only teamId
    training_data <- filtered_data %>%
      filter(teamId %in% train_teams) %>%
      dplyr::select(-teamId)

    # Testing data: keep all predictor columns, remove only teamId
    testing_data <- filtered_data %>%
      filter(teamId %in% test_teams) %>%
      dplyr::select(-teamId)
    
    # Random Forest
    set.seed(100 + k) # each fold reproducible but unique seed
    
    random_forest <- randomForest(Success ~ ., data = training_data, 
                                  ntree = 250, nodesize = 20, mtry = 5)
    
    # Class prediction
    y_hat_class <- predict(random_forest, testing_data, type = "class")
    
    # Player-level accuracy
    fold_accuracy <- confusionMatrix(y_hat_class, testing_data$Success, 
                                     positive = "c1")$overall["Accuracy"]
    
    # Predicted probability of Top-10 class
    y_hat_prob <- predict(random_forest, testing_data, type = "prob")[, "c1"]
    
    # Aggregate player probabilities to team level
    team_predictions <- data.frame(
      teamId = test_team_ids,
      actual = testing_data$Success,
      predicted_prob = y_hat_prob
    ) %>%
      group_by(teamId) %>%
      summarise(
        actual = first(actual),
        predicted_prob = mean(predicted_prob),
        .groups = "drop"
      ) %>%
      mutate(
        fold = k,
        min_appearances = min_appearances
      )
    
    all_team_predictions <- rbind(
      all_team_predictions,
      team_predictions
    )
    
    # Player-level ROC / AUC
    roc_obj <- roc(testing_data$Success, y_hat_prob, levels = c("c0", "c1"),
                 direction = "<")
    fold_auc <- as.numeric(auc(roc_obj))
    
   # Save player-level fold results
    fold_results <- rbind(
      fold_results,
      data.frame(
        fold = k,
        min_appearances = min_appearances,
        n_training_players = nrow(training_data),
        n_testing_players = nrow(testing_data),
        accuracy = as.numeric(fold_accuracy),
        auc = fold_auc
      )
    )
  }

  list(
    fold_results = fold_results,
    team_predictions = all_team_predictions
  )
}
```

```{r threshold-models}
# All appearance thresholds tested
appearances_thresholds <- c(0, 3, 10, 20, 25)

# Run grouped cross-validation once for each appearance threshold

threshold_models <- lapply(
  appearances_thresholds,
  function(threshold) {
    run_grouped_cv(
      player_data = player_data_ready,
      min_appearances = threshold,
      team_folds = team_folds
    )
  }
)

names(threshold_models) <- appearances_thresholds
```

## Appearance Threshold Analysis

I wanted to see what the effects of playing time (appearances_value) would
have on the model performance as players with low appearances may influence
the results. I created minimum appearance thresholds (0, 3, 10, 20, 25) that
will be evaluated in the grouped cross-validation procedure. 

Performance is evaluated at both the player and team level. Player metrics
measure how well the model can identify a Top 10 team player versus a Bottom 10
team player. Team metrics are calculated by averaging the out of fold predicted
probabilities of players within each club.
```{r player-threshold-results}
# run 5-fold cross validation all appearance thresholds
cv_results <- do.call(
  rbind,
  lapply(
    threshold_models,
    function(model) model$fold_results
  )
)

cv_summary <- cv_results %>%
  group_by(min_appearances) %>%
  summarise(
    mean_accuracy = mean(accuracy),
    sd_accuracy = sd(accuracy),
    mean_auc = mean(auc),
    sd_auc = sd(auc),
    .groups = "drop"
  )

cv_summary
```

```{r team-threshold-results}
team_threshold_results <- do.call(
  rbind,
  lapply(
    threshold_models,
    function(model) {

      team_predictions <- model$team_predictions %>%
        mutate(
          predicted_class = factor(
            ifelse(predicted_prob >= 0.5, "c1", "c0"),
            levels = c("c0", "c1")
          )
        )

      team_roc <- roc(
        team_predictions$actual,
        team_predictions$predicted_prob,
        levels = c("c0", "c1"),
        direction = "<"
      )

      data.frame(
        min_appearances = unique(team_predictions$min_appearances),
        team_accuracy = mean(
          team_predictions$predicted_class ==
            team_predictions$actual
        ),
        team_auc = as.numeric(auc(team_roc))
      )
    }
  )
)

team_threshold_results
```

```{r threshold-plots}
# Plot mean AUC across appearance thresholds
ggplot(cv_summary, aes(x = min_appearances, y = mean_auc)) +
  geom_line(linewidth = 1) + geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_auc - sd_auc,
      ymax = mean_auc + sd_auc
        ),
        width = 1) +
        scale_x_continuous(breaks = appearances_thresholds) +
  labs(
    title = "Model Performance by Minimum Appearance Threshold",
    subtitle = "5-fold cross validation grouped by team",
    x = "Minimum Player Appearances",
    y = "Mean AUC"
  ) + theme_minimal()

# Plot accuracy across appearance thresholds
ggplot(
  cv_summary,
  aes(x = min_appearances, y = mean_accuracy)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(
      ymin = mean_accuracy - sd_accuracy,
      ymax = mean_accuracy + sd_accuracy
    ),
    width = 1
  ) +
  scale_x_continuous(
    breaks = appearances_thresholds
  ) +
  scale_y_continuous(
    labels = scales::percent_format()
  ) +
  labs(
    title = "Classification Accuracy by Minimum Appearance Threshold",
    subtitle = "5-fold cross-validation grouped by team",
    x = "Minimum Player Appearances",
    y = "Mean Accuracy"
  ) +
  theme_minimal()
```

Filtering out players with low appearances did improve model performance. At
the player level, mean cross-validated AUC increased from 0.76 to 0.89 when
moving from no appearance restriction to 10+ appearance restriction. Higher
appearance thresholds did not provide any significant improvement to 
player level AUC, and did not consistently improve team-level performance. 

I chose the 10+ appearance threshold for my final analysis. It achieved 
approximately 81% mean player level accuracy and 0.89 mean AUC under team
grouped cross-validation, while aggregated out of fold predictions achieved
70% accuracy and 0.79 AUC across all 20 soccer teams. The 10+ appearance 
threshold allowed us to keep more observations with relatively the same
accuracy as the higher appearance thresholds.


## Final Random Forest Model

After selecting the 10+ appearance threshold, a final Random Forest is fit
using all the player still left after filtering. This model is fit to the 
full selected data set for feature interpretation. Predictive performance is
still reported using the team grouped cross-validation method rather than 
the in-sample performance of this final model.
```{r final-random-forest}
# 10+ appearance threshold 
final_data <- player_data_ready %>% filter(appearances_value >= 10) %>% 
              select(-teamId)

set.seed(123)

final_random_forest <- randomForest(Success ~ ., data = final_data, 
                                  ntree = 500, nodesize = 20, mtry = 5,
                                  importance = TRUE)

final_random_forest
```

```{r feature-importance-plot}
importance_matrix <- importance(final_random_forest)

importance_df <- data.frame(feature = rownames(importance_matrix), 
                            mean_decrease_accuracy = 
                              importance_matrix[, "MeanDecreaseAccuracy"],
                            mean_decrease_gini = 
                              importance_matrix[, "MeanDecreaseGini"]) %>%
  arrange(desc(mean_decrease_accuracy))

importance_df %>%
  slice_max(order_by = mean_decrease_accuracy, n = 10) %>%
  ggplot(
    aes(
      x = reorder(feature, mean_decrease_accuracy),
      y = mean_decrease_accuracy
    )
  ) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Random Forest Feature Importance",
    subtitle = "Players with 10+ appearances",
    x = NULL,
    y = "Mean Decrease Accuracy"
  ) +
  theme_minimal()
```

The strongest predictor in the final Random Forest is goalsConceded,
followed by appearances and goal assists. I wanted to dig into goalsConceded 
as it seems to be very closely related to match outcomes and consequently
league standings. 

These results indicate which variables the fitted Random forest used the most
when trying to distinguish a Top 10 team player from a Bottom 10 team player.

## Feature Robustness Analysis

I did additional analysis on the goalsConceded variable to see how dependent the
model is on this particular variable as it accounts for a large portion of 
the model's predictive signal. The variable does vary across players within the
same team, and logically is also correlated with appearances, so it is not
necessarily just a team level value.

Sensitivity models are evaluated using the same team grouped cross-validation
procedure, first I exclude goalsConceded and then exclude goalsConceded,
saves, and shotsFaced all together.
```{r goal-conceded-analysis}

final_data %>%
  summarise(
    n = n(),
    goals_conceded_zero = sum(goalsConceded_value == 0),
    goals_conceded_positive = sum(goalsConceded_value > 0),
    pct_zero = mean(goalsConceded_value == 0)
  )

final_data %>%
  summarise(
    saves_zero = sum(saves_value == 0),
    saves_positive = sum(saves_value > 0),
    shots_faced_zero = sum(shotsFaced_value == 0),
    shots_faced_positive = sum(shotsFaced_value > 0)
  )


final_data %>%
  select(
    goalsConceded_value,
    saves_value,
    shotsFaced_value
  ) %>%
  cor()

player %>%
  filter(appearances_value >= 10) %>%
  group_by(teamId) %>%
  summarise(
    n_players = n(),
    n_unique_goals_conceded = n_distinct(goalsConceded_value),
    min_goals_conceded = min(goalsConceded_value, na.rm = TRUE),
    max_goals_conceded = max(goalsConceded_value, na.rm = TRUE)
  )

cor(
  final_data$appearances_value,
  final_data$goalsConceded_value
)
```

```{r sensitivity-analysis}
# remove goalsConceded, saves_value, shotsFaced_value
player_data_sensitivity <- player_data_ready %>%
  select(
    -goalsConceded_value,
    -saves_value,
    -shotsFaced_value
  )

sensitivity_cv <- run_grouped_cv(
  player_data = player_data_sensitivity,
  min_appearances = 10,
  team_folds = team_folds
)

sensitivity_cv$fold_results

sensitivity_cv$fold_results %>%
  summarise(
    mean_accuracy = mean(accuracy),
    sd_accuracy = sd(accuracy),
    mean_auc = mean(auc),
    sd_auc = sd(auc)
  )

#remove goals_conceded
player_data_no_goals_conceded <- player_data_ready %>%
  select(-goalsConceded_value)

no_goals_conceded_cv <- run_grouped_cv(
  player_data = player_data_no_goals_conceded,
  min_appearances = 10,
  team_folds = team_folds
)

no_goals_conceded_cv$fold_results %>%
  summarise(
    mean_accuracy = mean(accuracy),
    sd_accuracy = sd(accuracy),
    mean_auc = mean(auc),
    sd_auc = sd(auc)
  )
```

Unfortunately removing goalsConceded greatlty reduced the models' predictive
performance, mean player level AUC fell from 0.89 to 0.54. The removal of 
goalsConceded, saves, and shotsFaced similarly dropped the mean AUC to 0.55.

The performance of the full model is strongly driven by the single variable
of goalsConceded. The model should be more useful in interpreting the
statistical differences between players on Top 10 and Bottom 10 teams rather
than forecasting future league standings.


## Principal Component Analysis

Principal Component Analysis (PCA) is used to examine variable redundancy and
structure of the 15 player performance variables used. Several variables
measure related aspects of involvement and performance, PCA transforms them into
unrelated componets to summarize any common patterns in the data.

PCA is used to explore rather than a predictive model. The same 10+ appearance
threshold will be used.
```{r pca-model}
#PCA: Player Performance
columns_pca <- c("totalGoals_value", 
                     "goalAssists_value", "yellowCards_value", 
                     "redCards_value", "appearances_value", 
                     "totalShots_value", "shotsOnTarget_value", 
                     "saves_value", "subIns_value", 
                     "foulsCommitted_value", "foulsSuffered_value",
                     "ownGoals_value", "offsides_value", 
                     "goalsConceded_value", "shotsFaced_value")

# Clean Data
pca_data <- player %>%
  filter(appearances_value >= 10) %>% 
  select(all_of(columns_pca)) %>% 
  na.omit()

pca_player <- prcomp(
  pca_data,
  center = TRUE,
  scale. = TRUE
)

summary(pca_player)
```

```{r pca-loadings}
loadings <- as.data.frame(pca_player$rotation)

loadings %>%
  select(PC1, PC2, PC3) %>%
  arrange(desc(abs(PC1)))

loadings %>%
  select(PC1, PC2, PC3) %>%
  arrange(desc(abs(PC2)))

loadings %>%
  select(PC1, PC2, PC3) %>%
  arrange(desc(abs(PC3)))

```

The first principal component (PC1) is dominated by totalShots, foulsSuffered, 
appearances, shotsOnTarget, fouls_committed, totalGoals, goalAssists. This
suggests that PC1 captures player involvments and statistical activity.

The second principal component (PC2) is dominated by saves, shotsFaced, 
goalsConceded. This PC captures defensive and goal keeper related statistics 
from other forms of player statistics. 

The third principal component is less defined and can not be clearly grouped
in terms of soccer. The main contributors are yellowCards, foulsCommitted,
shotsFaced, and totalGoals.

```{r variance-plot}
# Cumulative Variance Explained

pca_variance <- pca_player$sdev^2 / sum(pca_player$sdev^2)

pca_summary <- data.frame(
  component = seq_along(pca_variance),
  variance_explained = pca_variance,
  cumulative_variance = cumsum(pca_variance)
)

ggplot(
  pca_summary,
  aes(x = component, y = cumulative_variance)
) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(
    yintercept = 0.80,
    linetype = "dashed"
  ) +
  scale_x_continuous(breaks = 1:15) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  labs(
    title = "Cumulative Variance Explained by Principal Components",
    subtitle = "Player statistics for players with 10+ appearances",
    x = "Number of Principal Components",
    y = "Cumulative Variance Explained"
  ) +
  theme_minimal()
```
The first three principal components explain about 66% of the total variance in
the player statistics, the first five PCs explain almost 80%.

This confirms that there is redundancy present among the 15 original variables. 
The PCA results complement the Random Forest diagnostics by showing that
many player statistics capture overlapping dimensions of playing time, attacking
involvment, and defensive exposure.


## Final Team-Level Evaluation

For a final evaluation on if player level predictions translate to meaningful
team level separation, the out of fold predicted probabilities from the 10+
appearance model are averaged within each club. The probabilities are generated
while that player's entire team is held out of the training data, these
predictions are not influenced by training on players from the same team.

```{r final-team-evaluation}
# Retrieve out-of-fold team predictions for selected 10+ threshold
team_results_10 <- threshold_models[["10"]]$team_predictions %>%
  mutate(
    predicted_class = factor(
      ifelse(predicted_prob >= 0.5, "c1", "c0"),
      levels = c("c0", "c1")
    )
  )

# Team-level confusion matrix
team_cm_10 <- confusionMatrix(
  team_results_10$predicted_class,
  team_results_10$actual,
  positive = "c1"
)

# Team-level ROC / AUC
team_roc_10 <- roc(
  team_results_10$actual,
  team_results_10$predicted_prob,
  levels = c("c0", "c1"),
  direction = "<"
)

team_auc_10 <- as.numeric(auc(team_roc_10))

team_cm_10
team_auc_10
team_results_10 %>%
  arrange(desc(predicted_prob))
```

Using a 0.5 classification threshold, the aggregated out of fold predictors
correctly classify 70% of the 20 Premier League clubs as Top 10 or Bottom 10
finishes in the league standings. The team level AUC is 0.79, indicating that
aggregated probabilities provide moderate differences between the two groups.

The results should be interpreted alongside the sensitivity analysis. 
Although player level statistics contain information that can help distinguish
between Top 10 and Bottom 10 teams, a large portion of the Random Forest's
predictive performance relied heavily on the goalsConceded which is a
same-season statistic. Therefore this analysis classifies team performance 
characteristics instead of being used for a potential forecasting of
future league standings.

## Conclusion

This analysis developed a Random Forest framework for distinguishing players
on Top 10 and Bottom 10 Premier League Teams while accounting for shared
outcomes of players from the same club. Team grouped cross-validation prevented
club level information from leaking between training and testing sets, and 
filtering by 10+ appearances allowed a good median approach for data retention
and predictive performance. 

The final selected model achieved 0.89 mean player level AUC under grouped
cross-validation, which aggregated out of fold predictions achieved 0.79 AUC
and 70% classification across the 20 clubs. Feature diagnostics and sensitivity
analysis showed that most of this performance was attributed to the
goalsConceded variable, empathizing the importance of evaluating whether
highly predictive features are closely related to the outcome being classified.
PCA also showed significant redundancy between the original player statistics,
with five principal components explaining about 79% of their variance. 


